Chasing \rainbows{} with the Habitable Worlds Observatory

### 2023-03-14 src/tex/ms.tex
[showyourwork] overleaf sync

\cite{2001Natur.412..885F} noted the potential for time-resolved photometry to identify water oceans on the surface of an exoplanet, which was further explored by \citet{williams2004} and \citet{giados2006}. Rotational variability of an exoplanet's reflected light colours can be inverted into a rough map of the planet's surface, with dark regions being suggestive of oceans \citep{2009ApJ...700..915C}, but specular reflection, a.k.a. glint, remains the most direct means to infer liquids on the surface of other worlds \citep[e.g., methane lakes on Titan;][]{2010GeoRL..37.7104S}. \citet{lustig2019} explored mapping exoplanets from their ocean glint in polarized light.

@@

@@

%% MARKER FOR WHERE KIM LEFT OFF PLEASE LEAVE %%%

% Move to Clouds?
In practice, clouds complicate the detection of ocean glint because they obscure the underlying surface and are themselves non-Lambertian scatterers. \cite{Zugger_2010} simulated photometric and polarized phase variations for planets including the effects of partial cloud coverage and ocean waves. Although the presence of clouds and atmospheric Rayleigh scattering weaken the signal of an underlying ocean, they again find that oceans are detectable and distinguishable from other sources of polarization. \cite{2010ApJ...721L..67R} compared observations of the Earth's scattering phase curve to comprehensive radiative transfer simulations; they found that the photometric brightening of Earth at crescent phases is mostly due to forward scattering from clouds, with a relatively small contribution from ocean glint.
\citet{Zugger_2011} note that ocean glint is best observed in near infrared opacity windows because of the reduced Rayleigh scattering at long wavelengths. This has the added advantage of minimizing the latitude--albedo effect, which could otherwise be a false-positive for photometric brightening at crescent phases due to glint \citep{2012ApJ...752L...3C}. Most recently, \cite{2019A&A...626A.129T,2022A&A...664A.172T} simulated the photometric and polarimetric signatures of terrestrial planets with inhomogeneous clouds. The bottom line is that a direct imaging mission sensitive to the phase-dependent reflected light photometry and polarization of an  exoplanet could establish its habitability.

\textcolor{blue}{Science motivation below: Kim}\\
\textbf{Why Polarization curves worth looking at [Kim]}
%compare to features in nonpol (strength rainbow); current capabilties?

@@

\textbf{Why rainbow worth it [Kim]} %calc; retrieve species including for non terrestrials; droplet liklihood of liquid surface; lit review bailey et below

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/accessible_phase_angles.png}  
    \caption{
        The big kahuna
        \timmy{What are the dashed vs. solid lines?}
    }
    \label{fig:betaallofit}
    \script{plot_n_versus_phase.py}
\end{figure*}

\subsubsection{Eccentric randomly inclined orbits}
\label{sec:eccentric}
We repeat the Monte Carlo simulation from \cref{sec:circular} assuming the orbital eccentricity follows a beta distribution with shape parameters $a=0.867$ and $b=3.03$ \citep{2013MNRAS.434L..51K}. 
%
We randomly drew 1000 random orbits for each star and determined the sky-projected orbits following, e.g., \cite{2010exop.book...15M}. Eccentricities were drawn from the beta distribution and inclinations were uniform in $\cos i$. 
%
We then filtered on \IWA's of 1, 2, 3 and 4 times $\lambda / D$, before identifying which targets would be observable at the scattering phase angles where ocean glint, rainbows, or the Rayleigh peak occur. 
%
\Cref{fig:ball-o-yarn} shows a sample of the generated orbits with the filtered \IWA's. The dashed lines in Figure \ref{fig:betaallofit} show the cumulative distribution for the most extreme scattering phase angles accessible, when accounting for orbital eccentricity.  Kepler's second law ensures that eccentric planets spend more time at orbital separations greater than their semi-major axis, and hence are more likely to peak out from behind the coronagraph IWA and extreme orbital phases (albeit with worse contrast due to the greater star--planet separation).

@@

\timmy{I think so far, the MC code doesn't compute angle coverage. I came up with this quick and dirty sketch to help us get started:}

@@

\timmy{Good question --- maybe Matt or Max know how the data for this was generated? (It's not part of the showyourwork yet)}

>>>>

\subsection{Ocean Glint}
\cite{2001Natur.412..885F} noted the potential for time-resolved photometry to identify water oceans on the surface of an exoplanet, which was further explored by \citet{williams2004} and \citet{giados2006}. Rotational variability of an exoplanet's reflected light colours can be inverted into a rough map of the planet's surface, with dark regions being suggestive of oceans \citep{2009ApJ...700..915C}, but specular reflection, a.k.a. glint, remains the most direct means to infer liquids on the surface of other worlds \citep[e.g., methane lakes on Titan;][]{2010GeoRL..37.7104S}. \citet{lustig2019} explored mapping exoplanets from their ocean glint in polarized light.

@@

% NEEDS Paragraph on mapping connection to tectonics

% NEEDS KOPPARLA (prob refer to , BERDYUGINA, ROSSI? (cloud?), HU?, added... blend cloud hinderance into argument for context to lead into rainbows, mention Lietzow and jeremy there., maybe some more mention of the Earthshine observations Karalidi, Gordon, etc

@@

% Move to Clouds?
In practice, clouds complicate the detection of ocean glint because they obscure the underlying surface and can act as strong forward scatterers, mimicking ocean glint. \cite{2010ApJ...721L..67R} compared observations of the Earth's scattering phase curve to radiative transfer simulations; they found that the photometric brightening of Earth at crescent phases is mostly due to forward scattering from clouds, with a relatively small contribution from ocean glint.
\cite{Zugger_2010} simulated photometric and polarized phase variations for Earth-like planets including the effects of partial cloud coverage in suppressing the Rayleigh scattering and ocean glint polarization. Although the presence of clouds and atmospheric Rayleigh scattering weaken the signal of an underlying ocean, they again find that oceans are detectable and distinguishable from other sources of polarization. To suppress the effects of Rayleigh scattering, glint observations can be made at longer wavelengths
\citep{Zugger_2011}.

%%% KIM's BOOKMARK FOR HERLSELF, PLEASE DO NOT DELETE %%%

\subsection{Polarized Phase Curves}
%compare to features in nonpol (strength rainbow); current capabilties?, errors from 3 papers,

% add latitude albedo degeneracy break here \citep{2012ApJ...752L...3C}

%This has the added advantage of minimizing the latitude--albedo effect, which could otherwise be a false-positive for photometric brightening at crescent phases due to glint \citep{2012ApJ...752L...3C}.

 %\cite{2019A&A...626A.129T,2022A&A...664A.172T} simulated the photometric and polarimetric signatures of terrestrial planets with inhomogeneous clouds.

@@

\subsection{Rainbows}
%calc; retrieve species including for non terrestrials; droplet liklihood of liquid surface; lit review bailey et below

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/accessible_phase_angles.png}  
    \caption{
        The big kahuna
        \timmy{What are the dashed vs. solid lines?}
        \timmy{How did this plot aggregate the simulations for each planet? mean? median? min/max?
        }
    }
    \label{fig:betaallofit}
    \script{plot_n_versus_phase.py}
\end{figure*}

\subsubsection{Eccentric randomly inclined orbits}
\label{sec:eccentric}
We repeat the Monte Carlo simulation from \cref{sec:circular} assuming the orbital eccentricity follows a beta distribution with shape parameters $a=0.867$ and $b=3.03$ \citep{2013MNRAS.434L..51K}. 
%
We randomly drew 1000 random orbits for each star and determined the sky-projected orbits following, e.g., \cite{2010exop.book...15M}. Eccentricities were drawn from the beta distribution, and inclinations were uniform in $\cos i$. 
%
We then filtered on \IWA's of 1, 2, 3 and 4 times $\lambda / D$, before identifying which targets would be observable at the scattering phase angles where ocean glint, rainbows, or the Rayleigh peak occur. 
%
\Cref{fig:ball-o-yarn} shows a sample of the generated orbits with the filtered \IWA's. The dashed lines in Figure \ref{fig:betaallofit} show the cumulative distribution for the most extreme scattering phase angles accessible, when accounting for orbital eccentricity.  Kepler's second law ensures that eccentric planets spend more time at orbital separations greater than their semi-major axis, and hence are more likely to peak out from behind the coronagraph IWA and extreme orbital phases (albeit with worse contrast due to the greater star--planet separation).

@@

@@

\timmy{Good question --- maybe Matt or Max know how the data for this was generated? Matt: yes, it's all in showyourwork right now. I ran 1000 sims for each target with a random eccentric orbit, then pulled out the betamax and betamin angles for each of the 1000 realisations. Max's plots encode the completeness as a function of phase angle, showing that you get about 10 percent more phenomena. }
